---
title: "IPTW Causal Project"
output:
  pdf_document: default
  html_notebook: default
---
```{r}
library(tidyverse)
library(readxl)
library(pander)
```


IPTW estimand:

$\Psi(\mathbb{P}_0)$ = $\mathbb{E}_0$$[\frac{\mathbb{I}(A=1)}{\mathbb{P_0}(A=1|W)}Y]$ - $\mathbb{E}_0$$[\frac{\mathbb{I}(A=0)}{\mathbb{P_0}(A=0|W)}Y]$

```{r}
set.seed(252)
ObsData <- read.csv("slpexcov1517.csv")

ObsData <- ObsData %>% dplyr::select(-SEQN, -exminwk, -slphrs, -household, -income, -snoring,
                              -apnea, -bmicat, -smoke, -alcohol, -phq9)

ObsData <- ObsData %>% mutate(A = targetex) %>% mutate(Y = targetslp) %>% 
  dplyr::select(-targetex, -targetslp)

ObsData <- na.omit(ObsData)

names(ObsData)
```
```{r}
summary(ObsData)
```


# 1) Create the propensity scores

First fit the logistic regression model:
```{r}
fit<-glm(A~age+factor(raceeth)+factor(educ)+factor(marital)+bmi+waist+factor(depressed),
    family="binomial",data=ObsData)
```

Get propensity scores:
```{r}
prob.1W <- predict(fit, type= "response") #prediced probability of getting the exercise

prob.0W <- 1 - prob.1W #prediced probability of not getting the exercise
```

look at distribution of propensity scores:

$\hat{\mathbb{P}}(A=1|W_i)$
```{r}
summary(prob.1W) %>% pander
```
$\hat{\mathbb{P}}(A=0|W_i)$
```{r}
summary(prob.0W) %>% pander
```

# 2) Create the weights:
```{r}
wt1 <- as.numeric(ObsData$A==1)/prob.1W
wt0 <- as.numeric(ObsData$A==0)/prob.0W
```

Look at weights:

$\frac{\mathbb{I}(A_i=1)}{\hat{\mathbb{P}}(A=1|W_i)}$
```{r}
summary(wt1) %>% pander
```
$\frac{\mathbb{I}(A_i=0)}{\hat{\mathbb{P}}(A=0|W_i)}$
```{r}
summary(wt0) %>% pander
```
# 3) Estimate:
```{r}
IPTW<- mean( wt1*ObsData$Y) - mean( wt0*ObsData$Y)

IPTW

unadj<-mean(ObsData$Y[ObsData$A==1])-mean(ObsData$Y[ObsData$A==0])
unadj
```

# 4) Arbitrarily truncate weights at 10, to see what happens:

First see how many weights are greater than 10:
```{r}
sum(wt1>10)
sum(wt0>10)
```
```{r}
wt1.trunc<- wt1
wt1.trunc[ wt1.trunc>10] <-10
```

```{r}
wt0.trunc<- wt0
wt0.trunc[ wt0.trunc>10] <-10
```

IPTW with truncated weights at 10:
```{r}
mean(wt1.trunc*ObsData$Y) - mean( wt0.trunc*ObsData$Y)
```
What about truncated at 5?
```{r}
sum(wt1>5)
sum(wt0>5)
```
```{r}
wt1.trunc5<- wt1
wt1.trunc5[ wt1.trunc5>5] <-5
```

```{r}
wt0.trunc5<- wt0
wt0.trunc5[ wt0.trunc>5] <-5
```

IPTW with truncated weights at 5:
```{r}
mean(wt1.trunc5*ObsData$Y) - mean( wt0.trunc5*ObsData$Y)
```
# 5) Stabilized IPTW estimator - Modified Horwitz Thompson estimator
```{r}
mean( wt1*ObsData$Y)/mean( wt1) - mean( wt0*ObsData$Y)/mean( wt0)

```

